---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

# pwiser

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of pwiser is to make applying arbitrary functions across combinations of columns within `{dplyr}` easy. Currently, the only function is `pairwise()`, which applies a function to all pairs of columns.

`pairwise()` is meant to be used within `mutate()` and `summarise()` verbs. It is largely inspired by `dplyr::across()`. 

pwiser sprang from conversations on the Rstudio Community thread and related conversations: [pairwise() function for use within dplyr::mutate() and dplyr::summarise()](https://community.rstudio.com/t/pairwise-function-for-use-within-dplyr-mutate-and-dplyr-summarise/94684).

## Example within `summarise()`

```{r example}
library(dplyr)
library(pwiser)
library(palmerpenguins)

penguins <- na.omit(penguins)
```

Example testing within-group correlations across columns of interest:
```{r}
cor_p_value <- function(x, y){
  stats::cor.test(x, y)$p.value
}

penguins %>% 
  group_by(species) %>% 
  summarise(pairwise(contains("_mm"), cor_p_value, .is_commutative = TRUE),
            n = n())
```

Setting `.is_commutative = TRUE` can save time on redundant calculations. 

When using `pairwise()` within `summarise()` the function(s) we apply should have an output length of 1. (Though can still output list-columns e.g. could replace `cor_p_value` above with the lambda function `~list(stats::cor.test(.x, .y))` if we wanted to keep entire output from `cor.test()`).

<!-- You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/master/examples>. -->

## Example within `mutate()`

Example getting ratios and differences of different columns:
```{r}
penguins %>% 
  mutate(pairwise(contains("_mm"), 
                  list(ratio = `/`, difference = `-`),
                  .names = "features_{.fn}_{.col_x}_{.col_y}")) %>% 
  glimpse()
```

`.names` can be used to customize naming convention.

## Installation

Install from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("brshallo/pwiser")
```

## See Also

There are a variety of other tools in R set-up for doing tidy pairwise operations. [widyr](https://github.com/dgrtwo/widyr) and the excellent [corrr](https://github.com/tidymodels/corrr) package (which is within the `tidymodels` suite) offer solutions (primarily) for summarising contexts (`corrr::colpair_map()` is the closest comparison as it also supports arbitrary functions). `recipes::step_ratio()` and `recipes::step_interact()` can be used for making pairwise products or ratios in mutating contexts. I cataloged my tweeted on these approaches in the Appendix section of a post on [Tidy Pairwise Operations](https://www.bryanshalloway.com/2020/06/03/tidy-2-way-column-combinations/#tweets).

The novelty of `pwiser::pairwise()` is its integration in both mutating and summarising contexts.

## Computation Speed

*For problems with lots of data you should use more efficient approaches.*

Matrix operations (as opposed to using dataframes) are much more computationally efficient for problems involving combinations (which can get big very quickly). We've done nothing to optimize the computation of functions run through pwiser. 

For example, when calculating pearson correlations, `pairwise()` calculates the correlation *separately* for each pair, whereas something like `stats::cor()` (or `corrr::correlate()` which I believe calls `cor()` under the hood) does this in a single matrix operation.

```{r, warning=FALSE, message = FALSE}
library(modeldata)

data(cells)
cells_numeric <- select(cells, where(is.numeric))

dim(cells_numeric)
```

Let's do a speed test using the numeric columns from the `cells` dataset, which represents 56 columns (which means `r choose(56, 2)` pairwise combinations or `r 56*55` permutations).

```{r}
set.seed(123)

microbenchmark::microbenchmark(
  cor = cor(cells_numeric),
  correlate = corrr::correlate(cells_numeric),
  colpair_map = corrr::colpair_map(cells_numeric, cor),
  pairwise = summarise(cells_numeric, pairwise(where(is.numeric), cor, .is_commutative = TRUE)),
  times = 10L,
  unit = "ms")
```

The `stats::cor()` and `corrr::correlate()` approaches are many times faster than using `pairwise()`. However `pairwise()` still only takes roughly one fifth of a second to calculate 1540 correlations in this case (so on relatively constrained problems is still quite usable).

`pairwise()` seems to be faster than `corrr::colpair_map()` (a more apples-to-apples comparison as both handle arbitrary functions), though much of this speed difference goes away when `.is_commutative = FALSE`.

